---
title: "Grouped pipeline: testing across states"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped pipeline: testing across states}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(stateR)
library(dplyr)
library(tidyr)
library(purrr)
```

## Overview

In practice, brain state analyses involve not just computing metrics but testing
whether those metrics differ between groups or associate with continuous
variables — and doing so across every state simultaneously. This vignette shows
the full pipeline:

1. Compute `nest_fo()`, `nest_dwell()`, or `clusters_markov()`
2. Unnest and filter to one state (or transition) at a time
3. Apply a statistical model — demonstrated here with base R, and compatible
   with [`ptestR`](https://github.com/CoDe-Neuro/ptestR) for permutation tests

---

## Simulated data

We simulate 20 subjects (10 term, 10 preterm), each with 80 time points and
five possible states (0–4). States are drawn with unequal probabilities to
create realistic differences:

```{r sim-data}
set.seed(2024)

n_sub <- 20
n_tp  <- 80

# Preterm subjects spend more time in state 0, term in state 2
state_probs <- list(
  term    = c(0.15, 0.20, 0.35, 0.15, 0.15),
  preterm = c(0.35, 0.20, 0.15, 0.15, 0.15)
)

tbl <- purrr::map_dfr(seq_len(n_sub), function(i) {
  grp   <- ifelse(i <= 10, "term", "preterm")
  probs <- state_probs[[grp]]
  tibble::tibble(
    subject = sprintf("sub-%02d", i),
    group   = grp,
    pma     = rnorm(1, mean = ifelse(grp == "term", 40, 34), sd = 2),
    time    = seq_len(n_tp),
    state   = sample(0:4, n_tp, replace = TRUE, prob = probs)
  )
})

dplyr::count(tbl, group)
```

---

## Step 1 — Compute fractional occupancy

```{r fo}
fo <- nest_fo(
  tbl   = tbl,
  vars  = c("subject", "group", "pma"),
  foVar = "state"
)

fo
```

Each row is a state; the `data` list-column holds one row per subject with
their `perc` value and any covariates passed via `vars`.

---

## Step 2 — Test each state

Unnesting gives a flat data frame per state. We can map a model across all
states using `purrr::map()`. Here we use a simple linear model (`lm`) to
test whether fractional occupancy differs by group — swap in
`ptestR::grouped_perm_glm()` for a permutation test:

```{r fo-test}
fo_results <- fo %>%
  dplyr::mutate(
    model = purrr::map(data, ~ lm(perc ~ group + pma, data = .x)),
    tidy  = purrr::map(model, broom::tidy)
  ) %>%
  dplyr::select(cluster, tidy) %>%
  tidyr::unnest(tidy) %>%
  dplyr::filter(term == "groupterm") %>%
  dplyr::select(cluster, estimate, statistic, p.value) %>%
  dplyr::arrange(p.value)

fo_results
```

Apply FDR correction across all states:

```{r fdr}
fo_results %>%
  dplyr::mutate(p.fdr = p.adjust(p.value, method = "fdr"))
```

---

## Step 3 — Dwell time pipeline

The same pattern works for dwell time:

```{r dwell-test}
dwell <- nest_dwell(
  tbl    = tbl,
  vars   = c("subject", "group", "pma"),
  foVar  = "state",
  sortBy = "time"
)

dwell %>%
  dplyr::mutate(
    model = purrr::map(data, ~ lm(mean_dwell ~ group + pma, data = .x)),
    tidy  = purrr::map(model, broom::tidy)
  ) %>%
  dplyr::select(cluster, tidy) %>%
  tidyr::unnest(tidy) %>%
  dplyr::filter(term == "groupterm") %>%
  dplyr::select(cluster, estimate, statistic, p.value)
```

---

## Step 4 — Markov transition pipeline

For transitions, the nesting key is `tag` rather than `cluster`:

```{r markov-test}
trans <- clusters_markov(
  tbl      = tbl,
  vars     = c("subject", "group", "pma"),
  cVar     = "state",
  sortBy   = "time",
  groupBy  = "subject",
  remIntra = TRUE
)

trans_results <- trans %>%
  dplyr::mutate(
    model = purrr::map(data, ~ lm(nCount ~ group + pma, data = .x)),
    tidy  = purrr::map(model, broom::tidy)
  ) %>%
  dplyr::select(tag, tidy) %>%
  tidyr::unnest(tidy) %>%
  dplyr::filter(term == "groupterm") %>%
  dplyr::select(tag, estimate, statistic, p.value) %>%
  dplyr::arrange(p.value)

head(trans_results, 10)
```

---

## Using `ptestR` for permutation tests

In the dHCP neonatal study, `stateR` outputs were passed directly into
`ptestR` permutation tests with 10 000 permutations. The swap is one line:

```{r ptestR, eval = FALSE}
library(ptestR)

fo %>%
  dplyr::mutate(
    res = purrr::map(data, ~ grouped_perm_glm(
      tbl         = .x,
      formla      = perc ~ group + pma,
      var_to_perm = "perc",
      permNum     = 10000,
      seed        = 42
    ))
  ) %>%
  dplyr::select(cluster, res) %>%
  tidyr::unnest(res) %>%
  dplyr::filter(term == "groupterm") %>%
  dplyr::mutate(p.fdr = p.adjust(p.perm, method = "fdr"))
```

The same pattern applies to `nest_dwell()` (replace `perc` with
`mean_dwell`) and `clusters_markov()` (replace `perc` with `nCount` and
`cluster` with `tag`).

---

## Notes on `vars` and covariate carriage

Any column passed in `vars` is preserved in the `data` list-column and
therefore available inside every `map()` call. Covariates like `pma`,
`sex`, or `motion_outliers` should be included in `vars` when you plan to
use them in the downstream model — they do not need any special handling.

One caution: `vars` for `nest_fo()` and `nest_dwell()` defines the grouping
for the metric computation. If you include a continuous covariate that varies
*within* a subject (e.g. a time-varying measure), each unique combination of
all `vars` values will become its own group. For subject-level covariates (one
value per subject) this is not a problem.

---

## Further reading

- `vignette("getting-started")` — introduction to the three metrics
- `vignette("markov-transitions")` — transition matrix in depth
- [`ptestR`](https://github.com/CoDe-Neuro/ptestR) — permutation tests
