---
title: "Markov transition probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Markov transition probabilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(stateR)
library(dplyr)
library(tidyr)
```

## What is a Markov transition?

A first-order Markov chain models the brain state sequence as a memoryless
process: the probability of entering state $j$ at time $t+1$ depends only on
the current state $i$ at time $t$, not on any earlier history.

`clusters_markov()` estimates these transition probabilities empirically from
observed state sequences. For each subject (or group), it:

1. Lags the state sequence by one time point to create source–target pairs.
2. Counts every observed `source → target` transition.
3. Normalises by the total number of transitions *out of* each source state,
   giving a row-stochastic estimate of the transition matrix.

The result captures **which states the brain tends to move to and from**, and
**how probable each transition is** — information that fractional occupancy
alone cannot provide.

---

## Simulated data

```{r sim-data}
set.seed(99)

tbl <- tibble::tibble(
  subject = rep(paste0("sub-", 1:8), each = 60),
  group   = rep(c("term", "preterm"), each = 4 * 60),
  time    = rep(seq_len(60), 8),
  state   = sample(0:3, 8 * 60, replace = TRUE,
                   prob = c(0.35, 0.25, 0.25, 0.15))
)
```

---

## Computing transitions

```{r markov-basic}
trans <- clusters_markov(
  tbl      = tbl,
  vars     = c("subject", "group"),
  cVar     = "state",
  sortBy   = "time",
  groupBy  = "subject",
  remIntra = FALSE
)

trans
```

The output is nested by `tag` — one row per unique source–target pair. Unnest
to see the per-subject probabilities:

```{r unnest}
trans_long <- trans %>%
  tidyr::unnest(data)

trans_long %>%
  dplyr::select(tag, source, target, subject, group, nCount) %>%
  head(16)
```

---

## The transition matrix

To reconstruct the full $K \times K$ transition matrix averaged across
subjects, summarise `nCount` by `source` and `target`:

```{r transition-matrix}
trans_long %>%
  dplyr::group_by(source, target) %>%
  dplyr::summarise(mean_prob = mean(nCount), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from  = target,
    values_from = mean_prob,
    names_prefix = "to_"
  ) %>%
  dplyr::arrange(source)
```

Each row sums to 1 (within rounding error): given the brain is in state
`source`, the row entries are the probabilities of transitioning to each
possible `target` on the next time step.

---

## Removing self-transitions

By default `remIntra = FALSE`, so self-transitions (state $i$ → state $i$)
are included. These reflect dwell time in the transition matrix. If you want
to study only *genuine state changes*, set `remIntra = TRUE`:

```{r no-intra}
trans_change <- clusters_markov(
  tbl      = tbl,
  vars     = c("subject", "group"),
  cVar     = "state",
  sortBy   = "time",
  groupBy  = "subject",
  remIntra = TRUE
)

trans_change %>%
  tidyr::unnest(data) %>%
  dplyr::group_by(tag, source, target) %>%
  dplyr::summarise(mean_prob = mean(nCount), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_prob))
```

With `remIntra = TRUE`, rows no longer sum to 1 because the diagonal has been
removed and the remaining probabilities are re-normalised over inter-state
transitions only.

---

## The `tag` convention

Transition tags follow the `"<source>_<target>"` pattern. For a four-state
system (states 0–3), there are 16 possible tags including self-transitions, or
12 excluding them.

To work with a specific transition of interest:

```{r single-tag}
trans %>%
  tidyr::unnest(data) %>%
  dplyr::filter(tag == "1_0") %>%
  dplyr::select(subject, group, nCount)
```

---

## `vars` vs `groupBy`

Two arguments control grouping in `clusters_markov()`, and it is worth
understanding the distinction:

- **`groupBy`** — the variable used to *compute* transitions. Lag and pair
  operations are performed within each level of `groupBy`. This should
  always be the subject (or session) identifier, so that transitions are never
  computed *across* subjects.
- **`vars`** — the variables *carried through* to the output. Include
  everything you want available for downstream analysis: subject ID, group,
  covariates. `groupBy` should be a subset of (or equal to) `vars`.

```{r vars-vs-groupby}
# groupBy = "subject" ensures no cross-subject transitions are counted.
# vars includes "group" so group membership is available for testing.
clusters_markov(
  tbl      = tbl,
  vars     = c("subject", "group"),
  cVar     = "state",
  sortBy   = "time",
  groupBy  = "subject"
) %>%
  tidyr::unnest(data) %>%
  dplyr::filter(tag == "0_1") %>%
  dplyr::select(subject, group, nCount)
```

---

## Further reading

- `vignette("getting-started")` — fractional occupancy and dwell time
- `vignette("grouped-pipeline")` — applying permutation tests across all transitions
- `?clusters_markov`
